\bf Exponential Family Tensor Regression with Covariates on Multiple Modes

### 2020-08-27 tensor_regress/jasa-latexsource/tensor-jasa.tex
jasa

Many contemporary scientific and engineering studies collect multi-way array data, a.k.a.\ tensors, accompanied by additional covariates. One example is neuroimaging analysis~\cite{sun2017store,zhou2013tensor}, in which the brain connectivity networks are collected from a sample of individuals. Researchers are often interested in identifying connection edges that are affected by individual characteristics such as age, gender, and disease status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\cite{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent friendships. In addition, features on nodes and edges are often available, such as people's personality and demographic location. It is of keen scientific interest to identify the variation in the connection patterns (e.g., transitivity, community) that can be attributable to the node features.

{\bf Our contributions.} This paper presents a general treatment to these seemingly different problems. We formulate the learning task as a regression problem, with tensor observation serving as a response, and the node features and/or their interactions forming the predictor. Figure~\ref{fig:intro1}b illustrates the general set-up we consider. The regression approach allows the identification of variation in the data tensor that is explained by the covariates. Our model greatly improves the classical tensor regression~\cite{rabusseau2016low,yu2016learning} by incorporating covariates from multiple modes and the interactions thereof. The statistical convergence of our estimator is established, and we quantify the gain in predictive power.

A related contribution is that our method allows a broad range of tensor types, including continuous, count, and binary observations. While previous tensor regression methods~\cite{zhao2012higher,yu2016learning} are able to analyze Gaussian responses, none of them is suitable for exponential distribution family of tensors. We develop a generalized tensor regression framework, and as a by product, our models allows heteroscedasticity by relating the variance of tensor entry to its mean.  This flexibility is particularly important in practice, because social network, brain imaging, or gene expression datasets are often non-Gaussian.

@@

{\bf Related work.} Our work is closely related to but also clearly distinctive from several lines of previous work. The first is a class of \emph{unsupervised} tensor decomposition such as Tucker and CP decomposition~\cite{de2000multilinear,hong2018generalized,zhang2018tensor}. Tucker decomposition is an unsupervised method that finds a low-rank representation of a data tensor. In contrast, our model is a {\it supervised} tensor model that identifies the association between a data tensor and multiple covariates. The low-rank structure is determined jointly by the tensor response and matrix covariates. The second line of work studies the network-response model~\cite{rabusseau2016low, li2017parsimonious}. Earlier development of this model focuses mostly on binary data in the presence of dyadic covariates~\cite{hoff2005bilinear}. We will demonstrate the enhanced accuracy as the order of data grows, and establish the general theory for exponential family which is arguably better suited to various data types.

>>>>

Nowadays, multi-dimensional arrays, a.k.a.\ tensors, are collected with additional covariates on multiple modes in many modern scientific and engineering studies. One example is neuroimaging~\citep{sun2017store,zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent the friendships. Other covariates such as people’s demographic information and the closeness of a friendship are often available. In both examples, it is of keen scientific interest to identify the variation in the response (e.g., brain connectivity, social community patterns) that may be affected by additional covariates. These seemingly different problems can be formulated as a regression problem in the tensor sense.

Emerging instances of the non-Gaussian datasets also require the effective tools applicable to a wide range of data, such as continuous, binary and count data. Examples include the Political Social science data \citep{hu2015scalable} which is a count dataset, and the brain connectivity network data \citep{wang2019common} which is usually encoded as binary datasets. Easily treating these datasets as Gaussian-valued data will yield unstable estimations and lead the predictions fall outside the valid range. A number of previous methods have been proposed \citep{yu2016learning, llosa2018tensor, raskutti2015convex, zhao2012higher} to address the tensor regression problem in various forms (e.g.\ scalar-to-tensor regression, tensor-response regression) from distinct perspectives. However, these methods assume the Gaussianity in the model, treating the input as Gaussian valued entries. Novel methods for exponential family tensor regression are necessary this time.

@@

{\bf Related work.} Our work is closely related to but also clearly distinctive from several lines of previous work.

The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of data tensor. In contrast, our model can be viewed a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and covariates. The low-rank structure is determined jointly by the tensor response and matrix covariates.

The second line of work studies the scalar-to-tensor regression models, in which the response is a scalar and the \textit{predictor} is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to theirs because we treat the tensor as a \textit{response}. The distinct model settings yield distinct focuses and interpretations. Take neuroimaging analysis as an example. The scalar-to-tensor regression aims to study the changes of the prediction outcomes when the brain connectivity network varies, while tensor-response model focuses on understanding the variation in the brain connectivity that may attribute to the individual features such as age, gender and disease status. Since scalar-to-tensor models do not utilize all the information in the response, they are not optimal when there exist correlations among response variables. In contrast, our model does not assume the independence among response entries, which allow the model to capture more structural information.

The third line of work studies the tensor-on-tensor regression models, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, llosa2018tensor, gahrooei2020multiple}. Though our model actually tackles the similar problem as tensor-on-tensor regression, our work differs with previous work in several ways. First, previous methods mainly develop the model with Gaussian error. Their models are not suitable for general cases. Second, our study differs from previous work in the scope of results. \citep{gahrooei2020multiple} verifies the performance of estimation mainly via simulations, while our study provides a theoretical evidence on estimation. \citep{lock2018tensor, llosa2018tensor} perform solid inferences on parameter estimation, however, they do not assess the estimation and prediction errors, which are studied in our paper. Risk bounds are well studied in~\citep{raskutti2015convex} and our convergence rate coincides with their results in the order-3 low-rank case. Whereas, the convexity assumption of the regularization in their model yields a convex optimization problem. In contrast,  our model tackles a non-convex optimization problem, and employee distinct techniques.

{\bf Our contribution.} In this article, we present a general model and the associated theory for exponential family tensor regression. Exponential family tensor observation serves as the response, and the node features and/or the interactions form the matrix predictor. Figure~\ref{fig:intro1}b illustrates the general set-up we consider. Due to the generalized framework, our models allows heteroscedasticity by relating the variance of tensor entry to its mean. This ﬂexibility is particularly important in practice.  The regression approach allows the identiﬁcation of variation in the data tensor that is explained by the covariates. We utilize a low-rank constraint in the regression coeﬃcient to encourage the sharing among tensor entries. The statistical convergence of our estimator is established, and we quantify the gain in predictive power.
